---
title: "Distribution Conversions and NORTA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distribution Conversions and NORTA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=4)
library(dataforge)
dataforge_options(verbose=FALSE, plot=FALSE)
```

## The NORTA Approach

NORTA (Normal-to-Anything) works by:

1. Simulating correlated standard normal variables with the target correlation structure.
2. Transforming each variable through the inverse CDF of the desired marginal distribution.

This preserves the rank-order correlation structure (Spearman correlations match) while giving each variable the correct marginal distribution.

```{r norta-diagram, echo=FALSE, fig.height=3}
set.seed(42)
z1 <- rnorm(300); z2 <- rnorm_pre(z1, r = 0.7)
b1 <- norm2beta(z1, 2, 5); p1 <- norm2pois(z2, lambda = 4)
par(mfrow=c(1,3), mar=c(3,3,2,1))
plot(z1, z2, pch=16, cex=0.5, col="#3366AA80",
     main="Latent normals", xlab="Z1", ylab="Z2")
hist(b1, breaks=20, main="Beta marginal", xlab="", col="#66AADD")
hist(p1, breaks=10, main="Poisson marginal", xlab="", col="#88CC88")
```

---

## Individual Conversion Functions

Every `norm2X()` function takes a numeric vector (on the normal scale) and returns values from distribution X.  The matching `X2norm()` function inverts the transformation.

### Beta Distribution

```{r beta}
z  <- rnorm(500)
b  <- norm2beta(z, shape1 = 2, shape2 = 5)
summary(b)        # bounded in [0, 1]
# Round-trip
z2 <- beta2norm(b, shape1 = 2, shape2 = 5)
all.equal(z, z2, tolerance = 1e-10)
```

### Poisson Distribution

```{r pois}
p <- norm2pois(rnorm(500), lambda = 4)
table(p)[1:8]
```

### Binomial Distribution

```{r binom}
b10 <- norm2binom(rnorm(500), size = 10, prob = 0.3)
mean(b10)      # should be ~3
var(b10)       # should be ~2.1
```

### Gamma Distribution

```{r gamma}
g <- norm2gamma(rnorm(300), shape = 2, rate = 0.5)
summary(g)
```

### Uniform Distribution

```{r unif}
u <- norm2unif(rnorm(200), min = 18, max = 65)
range(u)
```

### Truncated Normal

```{r trunc}
t <- norm2trunc(rnorm(500), min = 0, max = 10, mu = 5, sd = 2)
range(t)        # strictly within [0, 10]
mean(t)         # should be ~5
```

---

## Likert Scale Simulation

`rlikert()` directly generates Likert responses; `norm2likert()` converts an existing normal variable.

```{r likert}
# Generate 7-point Likert responses
ratings <- rlikert(n = 200, items = 7, mu = 0, sd = 1)
table(ratings)

# Simulate biased (positive skew) ratings
pos_ratings <- norm2likert(rnorm(200, mean = 1), items = 7)
table(pos_ratings)
```

---

## Correlated Mixed Distributions with rmulti()

`rmulti()` is the all-in-one NORTA function. Specify a named list of distributions, a correlation structure, and sample size.

```{r rmulti-basic}
set.seed(99)
survey <- rmulti(
  n = 400,
  dist = list(
    age      = list(dist = "norm",   mu = 38, sd = 12),
    income   = list(dist = "gamma",  shape = 2, rate = 0.02),
    n_purch  = list(dist = "pois",   lambda = 6),
    rating   = list(dist = "likert", items = 5),
    loyal    = list(dist = "binom",  size = 1, prob = 0.6)
  ),
  r = 0.35
)
head(survey)
round(cor(survey), 2)
```

---

## Heterogeneous Correlation Structures

Pass a full correlation matrix or lower-triangle vector to `rmulti()`:

```{r rmulti-het}
# Correlation matrix: age & income highly correlated, others weakly
cm <- cormat(c(
  0.65,          # age ~ income
  0.10, 0.10,    # age ~ n_purch, income ~ n_purch
  0.20, 0.15, 0.05  # *~ rating pairs, n_purch ~ rating
), n = 4, var_names = c("age","income","n_purch","rating"))

df_het <- rmulti(
  n = 300,
  dist = list(
    age     = list(dist = "norm",   mu = 38, sd = 12),
    income  = list(dist = "gamma",  shape = 2, rate = 0.02),
    n_purch = list(dist = "pois",   lambda = 6),
    rating  = list(dist = "likert", items = 5)
  ),
  r = cm
)
round(cor(df_het), 2)
```

---

## Integrating with forge_design

Convert within-subject columns to non-normal distributions after simulation:

```{r integrate}
# Start with normal within-subject data
df <- forge_design(
  within = list(time = c("pre","post")),
  n = 100, mu = c(0, 1), sd = 1, r = 0.5
)

# Convert to Likert ratings (5-point)
df$pre_lik  <- norm2likert(df$pre,  items = 5)
df$post_lik <- norm2likert(df$post, items = 5)
table(df$pre_lik, df$post_lik)
```
