---
title: "Getting Started with dataforge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dataforge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(dataforge)
dataforge_options(verbose = FALSE, plot = FALSE)
```

## Overview

`dataforge` makes it easy to simulate realistic data for factorial and mixed experimental designs.  Whether you need a quick pilot dataset, want to run power analyses via simulation, or want to teach students about statistical concepts, `dataforge` provides a clean, pipe-friendly interface.

---

## 1. Simulating a Within-Subject Design

The main function is `forge_design()`.  Specify `within` factors as a named list, where each element is a named character vector of level labels.

```{r within-only}
library(dataforge)

# 3-level within-subject design (e.g., pre/mid/post)
df_w <- forge_design(
  within = list(time = c("pre", "mid", "post")),
  n      = 40,
  mu     = c(10, 12, 15),   # means at each time point
  sd     = 3,               # common SD
  r      = 0.6              # correlation among time points
)

head(df_w)
```

The returned object is a plain `data.frame` with a subject ID column and one column per within-subject cell.

---

## 2. Simulating a Between-Subject Design

For purely between-subjects designs, leave `within` at its default and specify `between`:

```{r between-only}
df_b <- forge_design(
  between = list(condition = c("control", "treatment")),
  n   = 50,    # 50 per group (100 total)
  mu  = c(10, 14),
  sd  = c(3, 3.5)
)

table(df_b$condition)
tapply(df_b$y, df_b$condition, mean)
```

---

## 3. Mixed (Split-Plot) Designs

Combine `within` and `between` for a mixed design.  Cell means are supplied in the order: **within levels vary fastest** (rows), then **between levels** (columns).

```{r mixed}
df_m <- forge_design(
  within  = list(time  = c("pre", "post")),
  between = list(group = c("control", "treatment")),
  n   = 40,                    # 40 per group
  mu  = data.frame(            # explicit matrix: rows=within, cols=between
    control   = c(10, 11),
    treatment = c(10, 16)
  ),
  sd  = 3,
  r   = 0.5
)

head(df_m)
```

---

## 4. Multi-Factor Designs

Both `within` and `between` accept multiple factors.

```{r multifactor}
df_2x3 <- forge_design(
  within  = list(
    time  = c("pre", "post"),
    cond  = c("A", "B")
  ),
  between = list(group = c("ctrl", "trt")),
  n  = 30,
  mu = c(5, 6, 7, 9,   # ctrl: pre_A, post_A, pre_B, post_B
         5, 7, 7, 12), # trt:  pre_A, post_A, pre_B, post_B
  sd = 2,
  r  = 0.4
)

dim(df_2x3)
names(df_2x3)
```

---

## 5. Plotting Designs

`plot_design()` produces a `ggplot2` violin-box plot, automatically detecting the design structure.

```{r plot, fig.height=4}
plot_design(df_m, x = "time", color = "group",
            title = "Pre/Post by Group")
```

For long-format data, specify `dv`:

```{r plot-long, fig.height=4}
df_long <- wide2long(df_m, within = "time", dv = "score")
plot_design(df_long, x = "time", color = "group", dv = "score")
```

---

## 6. Simulating Correlated Variables

`rnorm_forge()` generates multiple correlated normally-distributed variables directly, without needing a factorial design structure.

```{r rnorm-forge}
# 4 variables, all pairwise r = 0.5
cogscores <- rnorm_forge(
  n         = 200,
  vars      = 4,
  mu        = c(100, 80, 90, 70),
  sd        = c(15, 10, 12, 8),
  r         = 0.5,
  var_names = c("iq", "memory", "attention", "speed")
)

round(cor(cogscores), 2)
```

For heterogeneous correlations, supply a lower-triangle vector or full matrix:

```{r het-cor}
df_het <- rnorm_forge(
  n    = 300,
  vars = 3,
  r    = c(0.8, 0.2, 0.3),  # corr(1,2), corr(1,3), corr(2,3)
  var_names = c("x", "y", "z")
)
round(cor(df_het), 2)
```

---

## 7. Generating a Variable Correlated to an Existing One

`rnorm_pre()` creates a new vector with a specified correlation to one or more existing variables.

```{r rnorm-pre}
x <- rnorm(200, mean = 50, sd = 10)
y <- rnorm_pre(x, r = 0.75, mu = 100, sd = 15)
cor(x, y)
```

---

## 8. Simulating from Existing Data

`forge_df()` estimates the multivariate distribution of an existing dataset and generates new observations from it.

```{r forge-df}
# Simulate 30 iris flowers per species, preserving correlations
new_iris <- forge_df(iris, n = 30, between = "Species")
nrow(new_iris)

# Check means are similar
aggregate(Sepal.Length ~ Species, data = new_iris, FUN = mean)
aggregate(Sepal.Length ~ Species, data = iris,     FUN = mean)
```

---

## 9. Mixed Effects Design Builders

For more complex nested or cross-classified designs (e.g., students nested within schools, items crossed with subjects), use the pipe-friendly builder functions.

```{r mixed-effects}
library(magrittr)

set.seed(42)
school_data <- add_random(school = 6) %>%
  add_random(student = 20, .nested_in = "school") %>%
  add_between(.by = "school",
              type = c("public", "private"),
              .prob = c(4, 2)) %>%
  add_between(.by = "student",
              gender = c("F", "M", "NB"),
              .prob = c(0.49, 0.49, 0.02)) %>%
  add_ranef(school = 5, student = 10)

head(school_data, 10)
table(school_data$type, school_data$gender)
```

---

## 10. Non-Normal Distributions (NORTA)

`rmulti()` uses the NORTA (Normal-to-Anything) approach to simulate correlated data from mixed distributions.

```{r rmulti}
survey <- rmulti(
  n = 300,
  dist = list(
    age      = list(dist = "norm",   mu = 40,  sd = 12),
    n_items  = list(dist = "pois",   lambda = 8),
    rating   = list(dist = "likert", items = 7),
    accuracy = list(dist = "beta",   shape1 = 5, shape2 = 2)
  ),
  r = 0.3
)

str(survey)
round(cor(survey), 2)
```

---

## 11. Distribution Conversion Utilities

Individual `norm2X()` and `X2norm()` functions let you transform variables one at a time.

```{r dist-convert}
z <- rnorm(500)

# Convert to various distributions
beta_vals  <- norm2beta(z, shape1 = 2, shape2 = 5)
pois_vals  <- norm2pois(z, lambda = 4)
likert5    <- norm2likert(z, items = 5)
binom_vals <- norm2binom(z, size = 10, prob = 0.3)

summary(beta_vals)
table(likert5)
```

---

## 12. Reshaping Data

`wide2long()` and `long2wide()` handle within-subject data.

```{r reshape}
# Wide to long
df_wide <- forge_design(
  within = list(time = c("pre","mid","post")), n = 10,
  mu = c(5,7,9), sd = 1, r = 0.6
)
df_long <- wide2long(df_wide, within = "time", dv = "score")
head(df_long)

# Long to wide
df_back <- long2wide(df_long, id = "id", within = "time", dv = "score")
head(df_back)
```

---

## 13. Contrast Coding

Attach contrast-coded columns to a data frame for use in regression models.

```{r contrasts}
df_c <- data.frame(
  group = factor(rep(c("low","mid","high"), each = 30))
)

# Helmert coding (each level vs mean of previous levels)
df_c <- add_contrast(df_c, "group", contrast = "helmert")
head(df_c)

# Polynomial coding (for ordered factors)
df_p <- add_contrast(df_c[, "group", drop=FALSE], "group", contrast = "poly")
head(df_p)
```

---

## 14. Codebooks and Parameter Extraction

```{r codebook}
forge_codebook(df_m, output = "print")
```

```{r get-params}
get_params(df_m, between = "group")
```

---

## 15. Global Options

```{r options, eval=FALSE}
# Turn off auto-plotting and messages globally
dataforge_options(verbose = FALSE, plot = FALSE)

# Use long format by default
dataforge_options(long = TRUE)

# Change column name separator (default "_")
dataforge_options(sep = ".")

# Set a global seed
dataforge_options(seed = 2024)

# See current settings
dataforge_options()
```

---

## Session Info

```{r sessioninfo}
sessionInfo()
```
