---
title: "Getting Started with dataforge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dataforge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(dataforge)
set.seed(42)
```

## Overview

`dataforge` makes it easy to simulate structured datasets with precisely
specified statistical properties. Whether you need data for power analysis,
a reproducible teaching example, or unit tests for an analysis pipeline,
`dataforge` gives you a clean, pipe-friendly API.

```{r quick}
library(dataforge)
```

---

## 1. Correlated normal variables

The simplest entry point is `rnorm_multi()`, which generates a data frame
of correlated normal variables.

```{r rnorm-multi}
df <- rnorm_multi(
  n    = 200,
  vars = c("reaction_time", "accuracy", "fatigue"),
  mu   = c(350, 0.85, 3.0),
  sd   = c(50,  0.10, 0.8),
  r    = c(0.4, -0.3, -0.5)   # lower-triangle: rt-acc, rt-fat, acc-fat
)

round(cor(df), 2)
```

Set `empirical = TRUE` when you want the sample statistics to match the
targets *exactly* — useful for demonstrations:

```{r rnorm-empirical}
df2 <- rnorm_multi(100, vars = c("x","y"), mu = c(0,5),
                   sd = c(1, 2), r = 0.7, empirical = TRUE)
colMeans(df2)
apply(df2, 2, sd)
cor(df2)[1,2]
```

---

## 2. Factorial designs with `forge_design()`

`forge_design()` is the main workhorse for factorial simulation.
Specify `within` and/or `between` factor lists, then provide means and SDs
(either as scalars or structured vectors/data frames).

### Between-subjects only

```{r between}
df_between <- forge_design(
  between = list(drug = c("placebo", "low_dose", "high_dose")),
  n  = 40,
  mu = c(5.0, 5.8, 7.2),
  sd = 1.5,
  plot = FALSE
)
get_params(df_between, between = "drug", dv = "y")
```

### Within-subjects only

```{r within}
df_within <- forge_design(
  within = list(time = c("baseline", "week4", "week8", "week12")),
  n  = 50,
  mu = c(20, 18, 15, 13),
  sd = 4,
  r  = 0.6,    # common correlation between time points
  plot = FALSE
)
head(df_within[, 1:6])
```

### Mixed (between × within) design

```{r mixed}
df_mixed <- forge_design(
  within  = list(time = c("pre", "post")),
  between = list(condition = c("control", "treatment")),
  n  = 30,
  mu = data.frame(
    control   = c(50, 52),
    treatment = c(50, 62),
    row.names = c("pre","post")
  ),
  sd   = 10,
  r    = 0.6,
  long = TRUE,
  plot = FALSE
)
head(df_mixed)
```

Plotting the design is built in:

```{r plot, fig.cap="Violin plot of the mixed design"}
plot_design(df_mixed, geom = "violin")
```

---

## 3. Simulate from existing data with `forge_df()`

`forge_df()` estimates the empirical covariance structure of an existing
data frame and simulates new data that matches it.

```{r forge-df}
new_iris <- forge_df(iris, n = 100, between = "Species")

# Compare original and simulated means
original <- aggregate(Sepal.Length ~ Species, iris, mean)
simulated <- aggregate(Sepal.Length ~ Species, new_iris, mean)
data.frame(Species  = original$Species,
           original = round(original$Sepal.Length, 2),
           simulated = round(simulated$Sepal.Length, 2))
```

---

## 4. Building mixed effects structures

For cross-classified or nested designs, use the `add_*` pipeline:

```{r mixed-effects}
set.seed(1)
d <- add_random(subject = 30) |>
  add_within(.by = "subject", item = paste0("I", 1:10)) |>
  add_between(.by = "subject",
              group = c("control","treatment"),
              .prob = c(0.5, 0.5)) |>
  add_ranef(.by = "subject", subj_int = 1.5) |>
  add_ranef(.by = "item",    item_int  = 0.8)

# Add a residual and compute DV
d$residual <- rnorm(nrow(d), 0, 1)
d$y <- 5 + (d$group == "treatment") * 1.5 + d$subj_int + d$item_int + d$residual
head(d[, c("subject","item","group","y")])
```

### Nested random effects

```{r nested}
school_data <- add_random(school = 5) |>
  add_random(class = 4, .nested_in = "school") |>
  add_random(student = 20, .nested_in = "class")

cat("Total students:", nrow(school_data), "\n")  # 5*4*20 = 400
```

---

## 5. Contrast coding

`dataforge` provides a complete set of contrast coding helpers:

```{r contrasts}
d <- data.frame(
  condition = factor(rep(c("A","B","C"), each = 20)),
  y = c(rnorm(20, 5), rnorm(20, 7), rnorm(20, 9))
)

# Treatment coding (A is reference)
d_coded <- add_contrast(d, "condition", contrasts = "treatment", base = 1)
head(d_coded)

# Inspect the coding matrix directly
contr_code_helmert(c("none","low","medium","high"))
```

---

## 6. Non-normal distributions

`rmulti()` uses the NORTA method to simulate correlated non-normal data:

```{r rmulti}
df_mixed_dist <- rmulti(
  n      = 300,
  dist   = c("norm", "pois", "binom", "beta"),
  params = list(
    list(mu = 5, sd = 1),
    list(lambda = 3),
    list(size = 10, prob = 0.4),
    list(alpha = 2, beta = 5)
  ),
  r = 0.4,
  seed = 1
)
head(df_mixed_dist)
sapply(df_mixed_dist, summary)
```

### Likert-scale data

```{r likert}
# 4-item 5-point Likert scale, moderately correlated
survey <- rlikert(n = 200, items = 4, likert_n = 5,
                  mu = c(3.5, 3.2, 4.0, 2.8),
                  sd = 0.8,
                  r  = 0.5,
                  seed = 1)
head(survey)
sapply(survey, table)
```

---

## 7. Distribution conversions

Any normal vector can be converted to a target distribution via the
`norm2*` family:

```{r conversions}
x <- rnorm(500)

tibble_like <- data.frame(
  normal   = x,
  beta     = norm2beta(x, alpha = 2, beta = 8),
  poisson  = norm2pois(x, lambda = 5),
  uniform  = norm2unif(x, min = 1, max = 10),
  truncated = norm2trunc(x, min = 0, max = Inf)
)
sapply(tibble_like, function(v) round(range(v), 2))
```

---

## 8. Messy data

Introduce realistic data quality issues:

```{r messy}
df_clean <- forge_design(between = list(g = c("A","B")),
                         n = 60, mu = c(0, 1), sd = 1, plot = FALSE, seed = 1)

df_messy <- df_clean |>
  make_missing(cols = "y", prop = 0.10) |>
  add_outliers(cols = "y", prop = 0.03) |>
  add_duplicates(n_dups = 5)

cat("Original rows:", nrow(df_clean), "\n")
cat("Messy rows:   ", nrow(df_messy), "\n")
cat("NAs in y:     ", sum(is.na(df_messy$y)), "\n")
```

---

## 9. Reshaping

```{r reshape}
# Wide to long
df_wide <- forge_design(within = list(t = c("T1","T2","T3")), n = 20,
                        mu = c(5, 6, 7), sd = 1, r = .5,
                        plot = FALSE, seed = 1)
df_long <- wide2long(df_wide, within_cols = c("T1","T2","T3"),
                     names_to = "timepoint", values_to = "score")

# Long back to wide
df_w2 <- long2wide(df_long, id_cols = "id",
                   names_from = "timepoint", values_from = "score")
dim(df_wide); dim(df_long); dim(df_w2)
```

---

## 10. Global options

```{r options}
# Suppress auto-plotting globally
dataforge_options(plot = FALSE, verbose = FALSE)

# Change separator for compound column names
dataforge_options(sep = ".")

# Check current settings
dataforge_options()

# Reset
dataforge_options(plot = TRUE, verbose = TRUE, sep = "_")
```

---

## Session info

```{r session}
sessionInfo()
```
