---
title: "Mixed Effects Data Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed Effects Data Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4)
library(dataforge)
dataforge_options(plot = FALSE, verbose = FALSE)
set.seed(42)
```

## Introduction

This vignette demonstrates how to use `dataforge`'s pipe-friendly
`add_*` family to build up complex mixed effects data generating
processes (DGPs) — the kind you'd use to simulate data for lme4 models.

---

## 1. Subjects-only random intercept model

```{r simple-re}
set.seed(1)
d <- add_random(subject = 40) |>
  add_between(.by = "subject",
              condition = c("control", "treatment")) |>
  add_ranef(.by = "subject", u0 = 1.2) |>
  within({
    residual <- rnorm(length(subject), 0, 1)
    y <- 5 + 1.5 * (condition == "treatment") + u0 + residual
  })

head(d)
```

Fitting the model back:

```{r fit-simple, eval=requireNamespace("lme4", quietly=TRUE)}
library(lme4)
m <- lmer(y ~ condition + (1 | subject), data = d)
summary(m)$coefficients
```

---

## 2. Cross-classified design (subjects × items)

```{r cross-classified}
set.seed(2)
d_cc <- forge_mixed_cc(
  subj_n     = 30,
  item_n     = 20,
  grand_mean = 500,
  subj_sd    = 80,
  item_sd    = 40,
  error_sd   = 60
)
head(d_cc)

# Random effect variance components
cat("Subject SD:", round(sd(tapply(d_cc$subj_re, d_cc$subj, mean)), 1), "\n")
cat("Item SD:   ", round(sd(tapply(d_cc$item_re, d_cc$item, mean)), 1), "\n")
```

---

## 3. Nested design (students within classes within schools)

```{r nested}
set.seed(3)
school_data <- add_random(school = 8) |>
  add_random(class = 4, .nested_in = "school") |>
  add_random(student = sample(15:25, 32, replace = TRUE),
             .nested_in = "class") |>
  add_between(.by = "school",
              school_type = c("public", "private"),
              .prob = c(0.75, 0.25)) |>
  add_ranef(.by = "school", school_int = 5.0) |>
  add_ranef(.by = "class",  class_int  = 3.0) |>
  within({
    residual <- rnorm(length(student), 0, 8)
    score <- 65 +
             3 * (school_type == "private") +
             school_int + class_int + residual
  })

cat("Total students:", nrow(school_data), "\n")
cat("Mean score by school type:\n")
print(tapply(school_data$score, school_data$school_type, mean))
```

---

## 4. Random slopes model

```{r random-slopes}
set.seed(4)
# Correlated random intercept and slope
d_rs <- add_random(subject = 40) |>
  add_within(.by = "subject",
             trial = as.character(1:20)) |>
  add_between(.by = "subject",
              group = c("A", "B")) |>
  add_ranef(.by = "subject",
            u0 = 1.5,   # random intercept SD
            u1 = 0.4,   # random slope SD
            .cors = 0.3) |>  # intercept-slope correlation
  within({
    trial_num <- as.numeric(trial)
    residual  <- rnorm(length(subject), 0, 0.8)
    y <- 10 +
         0.5 * trial_num +
         1.0 * (group == "B") +
         u0 +
         u1 * trial_num +
         residual
  })

head(d_rs)
```

---

## 5. Longitudinal design with attrition

Build a full longitudinal dataset then introduce realistic dropout:

```{r attrition}
set.seed(5)
d_long <- add_random(subject = 100) |>
  add_within(.by = "subject",
             wave = c("w1","w2","w3","w4","w5")) |>
  add_between(.by = "subject",
              arm = c("placebo","active"),
              .prob = c(.5,.5)) |>
  add_ranef(.by = "subject", u0 = 2.0) |>
  within({
    wave_num <- as.numeric(factor(wave, levels=c("w1","w2","w3","w4","w5")))
    residual <- rnorm(length(subject), 0, 1.5)
    y <- 20 +
         (arm == "active") * wave_num * 0.8 +
         u0 + residual
  })

# Introduce 20% dropout increasing with wave
set.seed(6)
d_long$dropout_prob <- 0.04 * as.numeric(factor(d_long$wave))
d_long$dropped <- runif(nrow(d_long)) < d_long$dropout_prob
d_observed <- d_long[!d_long$dropped, ]

cat("Complete rows:", nrow(d_long), "\n")
cat("After dropout:", nrow(d_observed), "\n")
cat("Retention by wave:\n")
print(table(d_observed$wave))
```

---

## 6. Adding contrasts to a mixed design

After building a mixed effects dataset you can add contrast codes for
use in `lmer` formulas:

```{r contrasts-lmer}
set.seed(7)
d_c <- add_random(subject = 60) |>
  add_within(.by = "subject",
             stimulus = c("word","nonword","pseudoword")) |>
  add_between(.by = "subject", skill = c("low","high")) |>
  add_ranef(.by = "subject", u0 = 50) |>
  within({
    residual <- rnorm(length(subject), 0, 30)
    rt <- 600 +
          50 * (stimulus == "nonword") +
          30 * (stimulus == "pseudoword") -
          40 * (skill == "high") +
          u0 + residual
  })

# Add sum contrasts for stimulus
d_c <- add_contrast(d_c, "stimulus", contrasts = "sum")
# Add treatment contrast for skill
d_c <- add_contrast(d_c, "skill", contrasts = "treatment", base = "low")

head(d_c[, c("subject","stimulus","skill","rt",
             "nonword.sum","pseudoword.sum","high.vs.low")])
```

---

## 7. Complete DGP: 2x2 mixed model for power analysis

Here's a complete workflow to simulate data for a power analysis:

```{r power-analysis}
simulate_and_test <- function(seed, n_subj = 40, effect = 0.5) {
  set.seed(seed)
  d <- add_random(subject = n_subj) |>
    add_within(.by = "subject", cond = c("A","B")) |>
    add_between(.by = "subject", group = c("G1","G2")) |>
    add_ranef(.by = "subject", u0 = 1.0) |>
    within({
      residual <- rnorm(length(subject), 0, 1)
      y <- 0 +
           effect * (cond == "B") +
           effect * (group == "G2") +
           effect * (cond == "B" & group == "G2") +
           u0 + residual
    })
  # Simple t-test on the B - A difference within each subject
  diffs <- tapply(d$y[d$cond=="B"], d$subject[d$cond=="B"], mean) -
           tapply(d$y[d$cond=="A"], d$subject[d$cond=="A"], mean)
  t.test(diffs)$p.value
}

# Estimate power over 200 simulations
p_values <- sapply(1:200, simulate_and_test, n_subj = 40, effect = 0.5)
cat(sprintf("Estimated power (n=40, d=0.5): %.0f%%\n", mean(p_values < .05) * 100))
```

---

## Summary of `add_*` functions

| Function | Purpose |
|----------|---------|
| `add_random(data, name = n)` | Add a random factor with `n` levels |
| `add_random(..., .nested_in = col)` | Nest new factor within existing |
| `add_within(.by, ...)` | Fully cross a within-subject factor |
| `add_between(.by, ..., .prob)` | Randomly assign between-subject levels |
| `add_ranef(.by, name = sd)` | Add normally distributed random effects |
| `add_recode(col, new_col, ...)` | Recode factor levels |
| `forge_mixed_cc(...)` | Shortcut for cross-classified design |
