---
title: "Mixed Effects Design Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed Effects Design Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=4)
library(dataforge)
library(magrittr)
dataforge_options(verbose = FALSE, plot = FALSE)
```

## Building Complex Designs with Pipes

The `add_*` family of functions uses a pipe-friendly API to construct complex nested or crossed designs step by step.

```
add_random()   — Add grouping / random factors
add_within()   — Expand each unit across within-subject levels
add_between()  — Assign between-subject factor levels
add_ranef()    — Simulate and add random effects
add_recode()   — Recode a factor column
add_contrast() — Attach contrast codes
```

---

## Example 1: Students Nested within Schools

```{r nested}
set.seed(1)

school_design <- add_random(school = 10) %>%
  add_random(student = 25, .nested_in = "school") %>%
  add_between(.by = "school",
              type = c("public", "private"),
              .prob = c(7, 3)) %>%
  add_between(.by = "student",
              year  = c("Y1","Y2","Y3"),
              .prob = c(1, 1, 1)) %>%
  add_ranef(school = 8, student = 15)

nrow(school_design)
head(school_design, 10)
```

Add a continuous outcome using the random effects:

```{r outcome}
school_design$score <- with(school_design, {
  # Fixed effects: public schools score 5 pts lower than private
  50 +
  ifelse(type == "private", 5, 0) +
  school_ranef +
  student_ranef +
  rnorm(nrow(school_design), 0, 5)
})

aggregate(score ~ type, school_design, mean)
```

---

## Example 2: Items × Subjects (Cross-Classified)

Many psycholinguistics or perception designs cross subjects with stimulus items.

```{r crossed}
set.seed(2)

item_design <- add_random(subject = 30, item = 20) %>%
  add_between(.by = "subject",
              condition = c("A", "B")) %>%
  add_ranef(subject = 20, item = 30) %>%
  {
    .$rt <- 500 +
      ifelse(.$condition == "B", 50, 0) +
      .$subject_ranef +
      .$item_ranef +
      rnorm(nrow(.), 0, 25)
    .
  }

head(item_design)
aggregate(rt ~ condition, item_design, mean)
```

---

## Example 3: Repeated Measures within Crossed Groups

```{r within-crossed}
set.seed(3)

design <- add_random(subject = 40) %>%
  add_between(group = c("control","treatment")) %>%
  add_within(session = c("s1","s2","s3")) %>%
  add_ranef(subject = 10)

# Add a linearly growing DV with treatment effect
design$outcome <- with(design, {
  session_num <- as.integer(session)
  base        <- 100
  growth      <- 5 * session_num
  trt_effect  <- ifelse(group == "treatment", 10 * session_num, 0)
  base + growth + trt_effect + subject_ranef + rnorm(nrow(design), 0, 8)
})

aggregate(outcome ~ group + session, design, mean)
```

---

## Example 4: Recoding and Contrasts in the Pipeline

```{r recode-contrasts}
df <- add_random(participant = 60) %>%
  add_between(age_group = c("young", "middle", "older")) %>%
  add_contrast("age_group", contrast = "helmert") %>%
  add_recode("age_group",
             young = 1, middle = 2, older = 3,
             .new_col = "age_num")

head(df)
```

---

## Fitting a Mixed Model

The output of the builder pipeline is a standard data frame, compatible with `lme4` and other mixed modelling packages.

```{r lme4, eval=FALSE}
library(lme4)

model <- lmer(outcome ~ group * session + (1 | subject),
              data = design)
summary(model)
```
