---
title: "Simulate by Design: Factorial Experiments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate by Design: Factorial Experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4
)
library(dataforge)
dataforge_options(plot = FALSE)
set.seed(42)
```

## Introduction

This vignette walks through the full capabilities of `forge_design()` for
simulating factorial experiments — from simple one-factor designs through
complex multi-factor mixed designs with per-cell specifications.

---

## The `forge_design()` parameter system

`forge_design()` accepts five key statistical parameters:

| Parameter | Description |
|-----------|-------------|
| `within`  | Named list of within-subject factors and their levels |
| `between` | Named list of between-subject factors and their levels |
| `n`       | Sample size(s) per between-cell |
| `mu`      | Cell means |
| `sd`      | Cell standard deviations |
| `r`       | Correlation among within-subject measurements |

The `mu` and `sd` parameters accept:
- A **scalar** — the same value for all cells
- A **vector** — one value per within-level (recycled across between-groups)
- A **data frame** — rows are within-levels, columns are between-groups

---

## 1. One-way between-subjects design

```{r one-way}
df <- forge_design(
  between = list(dose = c("0mg", "10mg", "50mg", "100mg")),
  n  = 25,
  mu = c(5.0, 5.5, 6.8, 7.9),
  sd = 1.2
)

get_params(df, between = "dose", dv = "y")
```

```{r one-way-plot, fig.cap="One-way dose effect"}
plot_design(df, geom = "box")
```

---

## 2. Two-way between-subjects (2×2) design

```{r two-way}
df_2x2 <- forge_design(
  between = list(
    therapy  = c("CBT", "medication"),
    severity = c("mild", "severe")
  ),
  n  = 30,
  mu = c(65, 55, 70, 45),   # CBT-mild, CBT-severe, med-mild, med-severe
  sd = 10
)

get_params(df_2x2, between = c("therapy", "severity"), dv = "y")
```

```{r two-way-plot, fig.cap="Therapy × Severity interaction"}
plot_design(df_2x2, x = "severity", color = "therapy", geom = "box")
```

---

## 3. Repeated-measures (within-subjects) design

```{r rm}
# 4-week longitudinal study
df_rm <- forge_design(
  within = list(week = c("baseline", "wk2", "wk4", "wk6")),
  n  = 60,
  mu = c(100, 92, 85, 80),
  sd = 15,
  r  = 0.7     # high temporal autocorrelation
)

# Check correlations between time points
w_cols <- c("baseline", "wk2", "wk4", "wk6")
round(cor(df_rm[, w_cols]), 2)
```

```{r rm-long}
df_rm_long <- wide2long(df_rm, within_cols = w_cols,
                        names_to = "week", values_to = "score")
plot_design(df_rm_long, x = "week", geom = "line", dv = "score")
```

---

## 4. Mixed factorial design (between × within)

```{r mixed}
df_mixed <- forge_design(
  between = list(group = c("control", "treatment")),
  within  = list(time  = c("pre", "mid", "post")),
  n  = 40,
  mu = data.frame(
    control   = c(50, 51, 52),
    treatment = c(50, 55, 63),
    row.names = c("pre", "mid", "post")
  ),
  sd = 8,
  r  = 0.65,
  long = TRUE
)

head(df_mixed, 8)
```

```{r mixed-plot, fig.cap="Group × Time interaction"}
plot_design(df_mixed, x = "time", color = "group", geom = "line", dv = "y")
```

---

## 5. Three-factor design (2×2×3)

```{r three-factor}
df_3f <- forge_design(
  between = list(
    sex      = c("M", "F"),
    training = c("novice", "expert")
  ),
  within = list(task = c("easy", "medium", "hard")),
  n  = 20,
  # mu: 3 within-levels × 4 between-cells (MN, ME, FN, FE)
  mu = data.frame(
    M.novice  = c(90, 75, 55),
    M.expert  = c(95, 88, 80),
    F.novice  = c(88, 73, 52),
    F.expert  = c(93, 87, 81),
    row.names = c("easy","medium","hard")
  ),
  sd = 8, r = 0.5, long = TRUE
)

cat("Rows:", nrow(df_3f), "\n")  # 20*4*3 = 240
aggregate(y ~ sex + training + task, data = df_3f,
          FUN = function(x) round(mean(x), 1))
```

---

## 6. Unequal cell sizes

Pass a named vector to `n` to specify different sizes per between-cell:

```{r unequal}
df_unequal <- forge_design(
  between = list(region = c("north", "south", "east", "west")),
  n  = c(north = 50, south = 40, east = 60, west = 35),
  mu = c(5.1, 4.8, 5.5, 4.9),
  sd = 1.0
)
table(df_unequal$region)
```

---

## 7. Per-cell standard deviations (heteroscedasticity)

```{r hetero}
df_hetero <- forge_design(
  between = list(group = c("A", "B", "C")),
  n  = 50,
  mu = c(10, 10, 10),    # equal means
  sd = c(1,   3,   6)   # very different spreads
)
aggregate(y ~ group, data = df_hetero,
          FUN = function(x) round(sd(x), 2))
```

---

## 8. Specifying correlation structures

### Uniform correlation (scalar `r`)

```{r r-scalar}
df_r <- forge_design(within = list(item = paste0("I", 1:5)),
                     n = 80, mu = 0, sd = 1, r = 0.4)
round(cor(df_r[, paste0("I", 1:5)]), 2)
```

### Lower-triangle vector

```{r r-triangle}
# 3 items: r(I1,I2)=.6, r(I1,I3)=.3, r(I2,I3)=.5
df_rt <- forge_design(within = list(item = c("I1","I2","I3")),
                      n = 100, mu = 0, sd = 1,
                      r = c(.6, .3, .5))
round(cor(df_rt[, c("I1","I2","I3")]), 2)
```

### Full correlation matrix

```{r r-matrix}
R <- matrix(c(1.0, 0.8, 0.3,
              0.8, 1.0, 0.5,
              0.3, 0.5, 1.0), 3, 3)
df_rm2 <- forge_design(within = list(v = c("V1","V2","V3")),
                       n = 100, mu = 0, sd = 1, r = R)
round(cor(df_rm2[, c("V1","V2","V3")]), 2)
```

---

## 9. Checking and inspecting designs

```{r check}
# Print design summary before simulating
check_design(
  within  = list(time = c("T1","T2","T3")),
  between = list(group = c("A","B"), sex = c("M","F")),
  n = 20
)
```

```{r get-params}
df <- forge_design(between = list(group = c("A","B","C")),
                   n = 50, mu = c(5, 6, 7), sd = c(1, 1.5, 2))
get_params(df, between = "group", dv = "y")
```

---

## 10. Converting to long format

`forge_design(..., long = TRUE)` returns data in long format, but you can
also reshape after the fact:

```{r reshape}
df_wide <- forge_design(
  within  = list(emotion = c("happy","sad","angry")),
  between = list(age_group = c("young","old")),
  n = 25, mu = 0, sd = 1, r = .4
)

df_long <- wide2long(df_wide,
                     within_cols = c("happy","sad","angry"),
                     names_to    = "emotion",
                     values_to   = "rating")

head(df_long)
nrow(df_long)  # 50 subjects * 3 emotions = 150 rows
```

---

## Summary

| Design type | Key parameters |
|-------------|---------------|
| Between only | `between`, `n`, `mu`, `sd` |
| Within only  | `within`, `n`, `mu`, `sd`, `r` |
| Mixed        | `within`, `between`, `n`, `mu`, `sd`, `r` |
| Unequal `n`  | `n = c(g1 = n1, g2 = n2, ...)` |
| Per-cell SD  | `sd` as vector aligned with between-cells |
| Full cormat  | `r` as an `n_within × n_within` matrix |
