---
title: "Simulate by Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate by Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=4)
library(dataforge)
dataforge_options(verbose = FALSE, plot = FALSE)
```

## What is "Simulate by Design"?

`forge_design()` lets you specify a complete experimental design — within factors, between factors, sample sizes, means, SDs, and correlations — and produces a realistic data frame in one call. This is ideal for:

- **Power analysis** by simulation (run `rep = 1000` and analyse each dataset)
- **Teaching** about ANOVA, mixed models, and repeated measures
- **Pilot data** when real data collection hasn't started yet
- **Creating worked examples** for papers or tutorials

---

## Specifying Factor Levels

Factor levels can be simple character vectors, or named vectors where names are short codes and values are labels. The names are used for column names (wide format); the values are stored as factor levels.

```{r factor-levels}
# Simple character vector
within_simple <- list(time = c("pre", "post"))

# Named vector: codes → labels
within_named <- list(
  time = c(t1 = "Baseline", t2 = "Week 4", t3 = "Week 8")
)

df <- forge_design(within = within_named, n = 20,
                   mu = c(10, 12, 15), sd = 2, r = 0.6)
names(df)          # uses the codes (t1, t2, t3) for column names
```

---

## Specifying Cell Means

Cell means can be provided in several equivalent ways.

### Scalar — same mean everywhere

```{r mu-scalar}
df1 <- forge_design(within = list(t = c("a","b","c")), n = 30,
                    mu = 10, sd = 2, r = 0.3)
colMeans(df1[,c("a","b","c")])
```

### Vector — one value per cell (within varies fastest)

For a 2 (between: group) × 3 (within: time) design, order is:
`ctrl_pre, ctrl_mid, ctrl_post, trt_pre, trt_mid, trt_post`

```{r mu-vector}
df2 <- forge_design(
  within  = list(time = c("pre","mid","post")),
  between = list(group = c("ctrl","trt")),
  n   = 50,
  mu  = c(10, 11, 12,   # ctrl
          10, 13, 17),  # trt
  sd  = 2, r = 0.5
)
```

### Data frame — most explicit

Rows = within-level combinations, columns = between-level combinations:

```{r mu-df}
mu_df <- data.frame(
  ctrl = c(10, 11, 12),
  trt  = c(10, 13, 17),
  row.names = c("pre","mid","post")
)
df3 <- forge_design(
  within  = list(time = c("pre","mid","post")),
  between = list(group = c("ctrl","trt")),
  n = 50, mu = mu_df, sd = 2, r = 0.5
)
```

---

## Unequal Group Sizes

Supply a vector of `n` values (one per between-subject cell):

```{r unequal-n}
df_uneq <- forge_design(
  between = list(group = c("A","B","C")),
  n  = c(20, 40, 30),    # unequal Ns
  mu = c(10, 12, 11),
  sd = 2
)
table(df_uneq$group)
```

---

## Heterogeneous Correlations

Correlations can differ by between-subject cell using a named list:

```{r het-r}
df_hr <- forge_design(
  within  = list(t = c("pre","post")),
  between = list(group = c("low","high")),
  n   = 50,
  mu  = c(10, 11, 10, 14),
  sd  = 2,
  r   = list(low = 0.7, high = 0.2)   # different r per group
)
```

---

## Heterogeneous SDs

SDs can vary across cells just like means:

```{r het-sd}
sd_mat <- data.frame(ctrl = c(2, 2, 2), trt = c(2, 3, 4),
                     row.names = c("pre","mid","post"))
df_hs <- forge_design(
  within  = list(time = c("pre","mid","post")),
  between = list(group = c("ctrl","trt")),
  n = 40, mu = c(10,11,12,10,13,17), sd = sd_mat, r = 0.5
)
```

---

## Simulation Studies with rep

Use `rep` to generate many datasets at once:

```{r rep}
datasets <- forge_design(
  within  = list(t = c("pre","post")),
  between = list(g = c("ctrl","trt")),
  n = 30, mu = c(0,0,0,2), sd = 1, r = 0.5,
  rep = 100
)

length(datasets)  # 100 datasets

# Run a t-test on each and extract p-values
p_values <- vapply(datasets, function(d) {
  t.test(d$post[d$g == "trt"], d$post[d$g == "ctrl"])$p.value
}, numeric(1))

mean(p_values < 0.05)  # empirical power estimate
```

---

## Checking a Design Before Simulating

`check_design()` validates your specification without generating data:

```{r check-design}
check_design(
  within  = list(time = c("pre","mid","post")),
  between = list(group = c("A","B")),
  n = 50, mu = c(10,12,14,10,15,20), sd = 2, r = 0.6
)
```

---

## Getting Parameters Back

After simulating, `get_params()` computes observed means, SDs, and correlations:

```{r get-params}
df <- forge_design(
  within  = list(time = c("pre","post")),
  between = list(group = c("ctrl","trt")),
  n = 100, mu = c(10,11,10,15), sd = 3, r = 0.6
)
get_params(df, between = "group")
```
